---
title: "FinalExam"
author: "Jiaying Liu"
date: "December 13, 2018"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1
The table below summarize data from a clinical trial to compare response to two treatments (a test drug and a placebo).
1) State the null and alternative hypothesis to be tested;
In this dataset, we have aa 2 by 2 contingency table for response and nonresponse of placebo and test_drug. We want to compare response in two group, so we can actually test the response rate in two group 

Assume the response in placebo is $p1$;

Assume the respnse in Treatment is $P2$
```{r}
n1=64
n2=60
p1=16/64
p2=40/60
t=(p1-p2)/(sqrt(p1*(1-p1)/n1+p2*(1-p2)/n2))
pvalue=2*(1-pnorm(abs(t),mean=0,sd=1))
cbind(t,pvalue)
```


## 2

```{r}
m_old=41.52
m_new=51.48
n_old=32
n_new=30
sd_old=17.15
sd_new=11.01
sp=((n_old-1)*(sd_old)^2+(n_new-1)*(sd_new)^2)/(n_old+n_new-2)
t=(m_old-m_new)/(sqrt(sp)*sqrt(1/n_old+1/n_new))
pvalue=pt(t,df=(n_old+n_new-2))
cbind(t,pvalue)

```
```{r}
t=(m_old-m_new)/(sqrt((sd_old)^2/n_old+(sd_new)^2/n_new))
c=((sd_old)^2/n_old)/(((sd_old)^2/n_old)+((sd_new)^2/n_new))
df=((n_old-1)*(n_new-1))/((n_old-1)*(1-c)^2+(n_new-1)*c^2)
pvalue=pt(t,df)
cbind(t,pvalue)
```



##3
#(a)
#(b)
#(c)
```{r}
p1=0.8
p2=0.85
lam1=1
lam2=1.1

#1
m1=1/lam1
m2=1/lam2

s1=1/lam1^2
s2=1/lam2^2

#2
#m1=lam1
#m2=lam2

#s1=lam1^2
#s2=lam2^2

v1=p1*s1+p1*(1-p1)*m1^2
v2=p2*s2+p2*(1-p2)*m2^2

c1=p1*m1-p2*m2

n=(v1+v2)/(c1^2)*((1.96+1.282)^2)
n
```

## (d)


```{r}
p1=0.8
p2=0.8
lam1=1
lam2=1
n1=10000 # testing 10,000 times
t1err=0
for (i in 1:n1){
n=24981
z1=rexp(n,rate=lam1)
u1=rbinom(n,1,p1)
x=z1*u1

z2=rexp(n,rate=lam2)
u2=rbinom(n,1,p2)
y=z2*u2

   if (((t.test(x,y,mu=0,alternative = "two.sided",var.equal = FALSE))$p.value)<=0.05) (t1err=t1err+1) 
}
cat("Type I error rate in percentage is", (t1err/n1)*100,"%")
```
https://stats.stackexchange.com/questions/148526/how-to-simulate-type-i-error-and-type-ii-error
```{r}
p1=0.8
p2=0.85
lam1=1
lam2=1.1
m1=1/lam1
m2=1/lam2
c1=p1*m1-p2*m2
#v1=p1*s1+p1*(1-p1)*m1^2
#v2=p2*s2+p2*(1-p2)*m2^2
#se=sqrt(v1/24973+v2/24973)




n1=10000 # testing 10,000 times
t2err=0
for (i in 1:n1){
n=24981
z1=rexp(n,rate=lam1)
u1=rbinom(n,1,p1)
x=z1*u1

z2=rexp(n,rate=lam2)
u2=rbinom(n,1,p2)
y=z2*u2

if (((t.test(x,y, mu=0))$p.value)<=0.05) (t2err=t2err+1) 
}
cat("power is", (t2err/n1)*100,"%")


```

```{r}

```

```{r}
library(lme4)
MyData=read.csv("d:/final3.csv")
m2=lmer(different~Control_treatment+Visit_number+Control_treatment*Visit_number+(Visit_number |PATID),data=MyData)
summary(m2)
```

